Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-09-30 paper/paper.tex
some comments to be discussed

The prospect of modeling timbre perception necessarily exceeds the symbolic domain.
Instead, it involves a cognitive process which arises from the subjective experience of listening.
From an audio signal processing standpoint, the task of simulating this cognitive process amounts to the design of a multidimensional feature space wherein some distance function evaluates arbitrary pairs of stimuli.
Rather than to merely discriminate instruments as mutually exclusive categories, this distance function must reflect a consensus of quantitative judgments of acoustic dissimilarity, all other parameters --- \ie{} duration, pitch, and intensity --- being equal.

Behind the overarching challenge of coming up with a robust predictive challenge for human behavior, the main practical application of timbre similarity retrieval lies in the emerging topic of computer-assisted instrumentation.
Following the aesthetic tradition of spectralism in contemporary music creation, the composer queries the machine with an arbitrary audio signal.
The outcome of this is an instrumental audio sample, alongside textual attributes of playing technique.
In this context, the computer serves as a bridge from the auditory level to the symbolic level, allowing to approximate a potentially infinite realm of timbral sensations by an orchestra of priorly specified extent and technical range.
To maximize the perceptual fidelity of such an approximation, a query-by-example search engine in the auditory domain should summarize the vastest possible vocabulary of timbral templates.

In this paper, we propose a machine listening model which computes the dissimilarity in timbre between any two audio samples.
The main originality of our model is that it encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
A second originality of our model is that we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.

@@

On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags'', encoded as binary outputs within some predefined label space.
Because such tags often belong to the publicly available metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.
Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign  allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset, maintained by the Community for Open and Sustainable Music Information Research (COSMIR), comprises a vast corpus of X polyphonic music recordings spanning X instruments, as a derivative of the Free Music Archive (FMA) dataset.
Another example is the Medley-solos-DB dataset, which comprises X monophonic excerpts from 8 instruments, as a derivative of the MedleyDB dataset of multitrack music.
Over the past decade, the surge of these large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning techniques, and notably deep convolutional networks, in music instrument recognition.
We refer to X for a recent review of the state of the art in this domain.

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRF).

@@

To construct a dataset for exploring the perception of timbre, we conducted a survey on the perceived similarity of a set of sounds.
The set consists of $78$ recordings of different instrument--PT (IPT) pairs, each played at the same pitch (C4) and nuance \textit{mezzo-forte}. For more information on the dataset considered in this study, see Section \ref{sec:dataset}.
In the survey, the subject is asked to organize dots on a two-dimensional place in a free sorting task.
Each dot corresponded to a different IPT recording and is moved by the user dragging them with the mouse.
In addition to spatially arranging the dots, the user is also asked to assign one of $20$ colors to each dot, providing a simple clustering of the sounds.
While both the location and color of each dot was recorded, only the color is used in our subsequent analysis---the spatial arrangement is only used to facilitate the user's ability to organize the sounds.
More information on the data collection process is available in  Section \ref{sec:judgments}.

@@

To evaluate the ability of the scattering transform to characterize the timbre of an audio signal, we consider the task of classifying musical notes from a wide variety of instruments played according to a set of extended instrumental playing techniques (IPTs).
Increasing the variability in IPTs allows us to explore a wider range of timbre, providing a more robust validation of the proposed model for modeling timbre.
The metric used is the \emph{precision at rank $k$} ($\patk$), which is calculated by computing the $k$ nearest neighbors in the feature space and recording the proportion of the neighbors that are in the same class.
In the following, we set $k = 5$.
The classes were obtained from the consensus clustering of human similarity judgments described in the previous section.
The $\patk$ therefore provides a measure of consistency between the timbre similarity measures obtained by our model and those of the survey subjects.

Concatenating the first- and second-order separable time and frequency scattering coefficients, we obtain a feature vector of dimension $???$ for a sound recording of duration $???$.
Computing nearest neighbors with the standard Euclidean measure gives a $\patk$ of $0.870 \pm 0.058$, where the first value indicates the mean and the second the standard deviation.
As a result, this simple transformation of the original signal results in a representation that lends itself naturally to discriminating timbre.
Note that this result is obtained by simply looking at the raw feature vectors---no learning is performed to adapt the transform towards discriminating timbre.
Replacing the separable second-order coefficients with the joint second-order coefficients boosts the $\patk$ to $???$.
This is evidence that accurately characterizing the joint time--frequency structure of the sound is important capturing its timbre.

We performed the same evaluation with a standard MFCC representation, obtaining a $\patk$ of $0.851 \pm 0.062$.
As a result, we conclude that a Euclidean distance over scattering transforms (separable or joint) yield little additional perceptual information compared to standard spectral descriptors.

@@

To improve our measure, we need to replace the standard Euclidean distance on the raw scattering coefficients by something better adapted to discriminating timbre.
Equivalently, we can apply further transformations to the scattering coefficients and compute the Euclidean distance on the result.
Very flexible and powerful methods can be brought to bear on this problem, such as deep neural networks, but with their flexibility comes an increase in complexity.
This increased complexity of the model reduces its usefulness with respect to interpretability.

For this reason, we restrict ourselves to a simple linear transform of the scattering coefficients.
Specifically, we construct a positive semidefinite matrix $L$ that we apply to the scattering coefficients $x$ so that the Euclidean distance between $Lx$ and $Ly$ best approximates the consensus clustering obtained above.
We expect the resulting representation $Lx$ to provide a more accurate model for timbre perception.

We now use the consensus clustering to reweight the similarity between feature vectors (either MFCCs or scattering transforms).
For this, we use the \emph{large margin nearest neighbor} (LMNN) metric learning algorithm \cite{weinberger2006distance, weinberger2009distance}.
This approach constructs a positive semi-definite weighting matrix $L$ such that the distance $\|Lx - Ly\|$ best classifies a set of points into a fixed clustering assignment.
Running LMNN on MFCCs with the consensus clustering obtained above increases $\patk$ slightly to $0.862 \pm 0.059$.
Consequently, there seems to be little additional timbre information in the MFCCs compared to that given by the standard Euclidean distance.
We contrast this with LMNN applied to the separable scattering coefficients, which yields a $\patk$ of $0.948 \pm 0.033$, which is a significant increase.
For the joint scattering coefficients, we obtain a $\patk$ of $???$.
We therefore conclude that a simple reweighting of the scattering transform yields a more appropriate timbral similarity measure.

@@

Considering the separable scattering gives a performance of $93\% \pm 4.2$.

@@

The success of the above system in identifying timbral similarities has immediate applications in browsing music databases.
These are typically organized based on instrumental and PT taxonomies, with additional keywords offering a more flexible organization.
Accessing the sounds in these databases therefore requires some knowledge of the taxonomy and keywords used.
Furthermore, the user needs to have some idea of the particular IPT they are searching for.

@@

The computational model proposed in this work is well suited to such a query-by-example task.
We have shown that it is able to approximate well the timbral judgments of a wide range of subjects included in our study.
Not only that, but the system can be easily retrained to approximate an individual user's timbre perception by having that user perform the clustering task on a reduced set of $78$ IPTs and running the LMNN training step on those clustering assignments.
This model can then be applied to new data, or, alternatively, be retrained with these new examples if the existing model proves unsatisfactory.

We shall note, however, that the current model has several drawbacks.
First, it is only applied to instrumental sounds.
While this has the advantage of simplifying the intepretation of the results, the range of timbre under consideration is necessarily limited (although less restricted than only considering \emph{ordinario} PTs).
This also makes applications such as \emph{query-by-humming} difficult, since we cannot guarantee that the timbral similarity measure is accurate for such sounds.

>>>>

The prospect of modeling timbre perception thus necessarily exceeds the symbolic domain.
Instead, it involves a cognitive process which arises from the subjective experience of listening.
From an audio signal processing standpoint, the task of simulating this cognitive process amounts to the design of a multidimensional feature space wherein some distance function evaluates arbitrary pairs of stimuli.
Rather than to merely discriminate instruments as mutually exclusive categories, this distance function must reflect a consensus of quantitative judgments of acoustic dissimilarity, all other parameters --- \ie{} duration, pitch, and intensity --- being equal.

Behind the overarching challenge of coming up with a robust predictive challenge for human behavior, the main practical application of timbre similarity retrieval lies in the emerging topic of computer-assisted instrumentation \ml{vl : cite esling and others ?}.
Following the aesthetic tradition of spectralism in contemporary music creation, the composer queries the machine with an arbitrary audio signal.
The outcome of this is an instrumental audio sample, alongside textual attributes of playing technique.
In this context, the computer serves as a bridge from the auditory level to the symbolic level, allowing to approximate a potentially infinite realm of timbral sensations by an orchestra of priorly specified extent and technical range.
To maximize the perceptual fidelity of such an approximation, a query-by-example search engine in the auditory domain should summarize the vastest possible vocabulary of timbral templates.

In this paper, we propose a machine listening model which computes the dissimilarity in timbre between any two audio samples.

\ml{decrire les contributions signal ici}

The main originality of our model is that it encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production \ml{vl cite dlfm ?}.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
A second originality of our model is that we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.

@@

On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem  \ml{vl cite}.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags'', encoded as binary outputs within some predefined label space.
Because such tags often belong to the publicly available metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.
Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign  allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset, maintained by the Community for Open and Sustainable Music Information Research (COSMIR), comprises a vast corpus of X polyphonic music recordings spanning X instruments, as a derivative of the Free Music Archive (FMA) dataset.
Another example is the Medley-solos-DB dataset, which comprises X monophonic excerpts from 8 instruments, as a derivative of the MedleyDB dataset of multitrack music.
Over the past decade, the surge of these large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning techniques, and notably deep convolutional networks, in music instrument recognition.
We refer to X for a recent review of the state of the art in this domain.

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \ml{vl cite}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRF) \ml{vl citer, dire que c'est du coup une base solide de représentation et que le modèle mathamtique que l'on utilisera ici est quasi equivalent}.

\ml{parler des travaux de Mounia}

@@

To construct a dataset for exploring the perception of timbre, we conducted a survey on the perceived similarity of a set of sounds.
The set consists of $78$ recordings of different instrument--PT (IPT) pairs, each played at the same pitch (C4) and nuance \textit{mezzo-forte}. For more information on the dataset considered in this study, see Section \ref{sec:dataset}.
In the survey, the subject is asked to organize dots on a two-dimensional place in a free sorting task.
Each dot corresponded to a different IPT recording and is moved by the user dragging them with the mouse.
In addition to spatially arranging the dots, the user is also asked to assign one of $20$ colors to each dot, providing a simple information of clustering of the sounds.
While both the location and color of each dot was recorded, only the color is used in our subsequent analysis---the spatial arrangement is only used to facilitate the user's ability to organize the sounds.
More information on the data collection process is available in  Section \ref{sec:judgments}.

@@

@@

The standard Euclidean distance on the raw scattering coefficients  provides some sort universal representation of timbre dissimilarity. Considering some sort of learning to apply further transformations to the scattering coefficients and compute the Euclidean distance on the result.

Due to the size of corpus, we restrict ourselves to a simple linear transform of the scattering coefficients. More flexible and powerful methods could be brought to bear on this problem, such as deep neural networks, but with their flexibility comes an increase of the needed size of training data.

Specifically, we construct a positive semidefinite matrix $L$ that we apply to the scattering coefficients $x$ so that the Euclidean distance between $Lx$ and $Ly$ best approximates the consensus clustering obtained above.
We expect the resulting representation $Lx$ to provide a more accurate model for timbre perception.

We now use the consensus clustering to reweight the similarity between feature vectors (either MFCCs or scattering transforms).
For this, we use the \emph{large margin nearest neighbor} (LMNN) metric learning algorithm \cite{weinberger2006distance, weinberger2009distance}.
This approach constructs a positive semi-definite weighting matrix $L$ such that the distance $\|Lx - Ly\|$ best classifies a set of points into a fixed clustering assignment.

@@

Considering the separable scattering gives a performance of $93\% \pm 4.2$. \ml{ ja write details on what is lost here ?}

@@

\ml{ parler des contributions signal ici}

The success of the above system in identifying timbral similarities has immediate applications in browsing music databases.
These are typically organized based on instrumental and playing techniques taxonomies, with additional keywords offering a more flexible organization.
Accessing the sounds in these databases therefore requires some knowledge of the taxonomy and keywords used.
Furthermore, the user needs to have some idea of the particular IPT they are searching for.

@@

The computational model proposed in this work is well suited to such a query-by-example task.
We have shown that it is able to approximate well the timbral judgments of a wide range of subjects included in our study.
Not only that, but the system can be easily retrained to approximate an individual user's timbre perception by having that user perform the clustering task on the reduced set of $78$ IPTs and running the LMNN training step on those clustering assignments.
This model can then be applied to new data, or, alternatively, be retrained with these new examples if the existing model proves unsatisfactory.

We shall note, however, that the current model has several drawbacks.
First, it is only applied to instrumental sounds.
While this has the advantage of simplifying the interpretation of the results, the range of timbre under consideration is necessarily limited (although less restricted than only considering \emph{ordinario} PTs).
This also makes applications such as \emph{query-by-humming} difficult, since we cannot guarantee that the timbral similarity measure is accurate for such sounds.

\ml{vl je reecrira le par du dessus en positif fture work}
